---
title: "DC Mortgage Data 2017"
author: "Andy Green"
date: "9/25/2020"
output: pdf_document
---

## Section 1: Set-up Tasks

```{r}
## Importing packages
library(tidyverse)
library(scales)
```

```{r}
## Read in the data
data <- read.csv("Data_Inputs/hmda_2017_dc_all-records_labels.csv", stringsAsFactors = FALSE)

## Convert to tibble
data <- as_tibble(data)

## Check it out
data
```

```{r}
## Narrowing down to just columns I care about
data <- data %>%
    select(respondent_id,property_type_name,loan_purpose_name,owner_occupancy_name,loan_amount_000s,preapproval_name,
           action_taken_name,census_tract_number,applicant_ethnicity_name,co_applicant_ethnicity_name,applicant_race_name_1,
           applicant_race_name_2,co_applicant_race_name_1,co_applicant_race_name_2,applicant_sex_name,co_applicant_sex_name,
           applicant_income_000s,denial_reason_name_1,denial_reason_name_2,denial_reason_name_3,lien_status_name,population,
           minority_population,tract_to_msamd_income,number_of_owner_occupied_units,number_of_1_to_4_family_units)

data
```

## Section 2: Exploratory Data Analysis and Clean-up

```{r}
## Making a scatter plot of income and loan amounts
plot1 <- ggplot(data = data, aes(x = applicant_income_000s, y = loan_amount_000s)) +
    geom_point(alpha = 0.4, color = "Blue") +
    ggtitle("Applicant Income and Loan Amount\n") +
    xlab("\nApplicant Income (Thousands)") +
    ylab("Loan Amount (Thousands)\n")

plot1 + scale_x_continuous(labels = comma) + scale_y_continuous(labels = comma)
```

```{r}
## Remove outliers > $50M
data <- data %>%
    filter(loan_amount_000s < 50000)

data
## note that we're now at 30,906 observations rather than 30,927
```

```{r}
## Repeat earlier graph w/o outliers
plot2 <- ggplot(data = data, aes(x = applicant_income_000s, y = loan_amount_000s)) +
    geom_point(alpha = 0.4, color = "Blue") +
    ylim(0,7500) +
    ggtitle("Applicant Income and Loan Amount\n") +
    xlab("\nApplicant Income (Thousands)") +
    ylab("Loan Amount (Thousands)\n")

plot2
```


```{r}
## Summarize loan amount by property type
property_type_1 <- data %>%
  group_by(property_type_name) %>%
  summarize(mean_loan_amount_000s = round(mean(loan_amount_000s)))

## Count of property types
property_type_2 <- data %>%
  group_by(property_type_name) %>%
  count()

## Merging together
property_type <- merge(property_type_1,property_type_2)
property_type$property_type_name[property_type$property_type_name == "One-to-four family dwelling (other than manufactured housing)"] <- "1-4 family dwelling"
property_type
```

```{r}
## Histogram
ggplot(data = property_type, aes(x = property_type_name, y = n)) + 
    geom_col(fill = "skyblue3") +
    geom_text(aes(label = n), size = 4, vjust= -0.5) +
    ylim(0,32000) +
    ggtitle("Number of Loans by Property Type\n") +
    xlab("\nProperty Type") +
    ylab("Number of Loans\n")
```


```{r}
## Making a simple bar chart to show disparities
ggplot(data = property_type, aes(x = property_type_name, y = mean_loan_amount_000s)) +
    geom_col(fill = "skyblue3") +
    geom_text(aes(label = mean_loan_amount_000s), size = 3.5, vjust= -0.5) +
    ylim(0,5000) +
    ggtitle("Average Loan Amount by Property Type\n") +
    xlab("\nProperty Type") +
    ylab("Average Loan Amount (Thousands)\n")
```

```{r}
## Dropping to just 1-4 family dwellings
data <- data %>%
    filter(property_type_name == "One-to-four family dwelling (other than manufactured housing)")

data
## note that we're now at 30,598 rows
```

```{r}
## Summarize loan amount by owner occupancy
owner_occupancy_1 <- data %>%
  group_by(owner_occupancy_name) %>%
  summarize(mean_loan_amount_000s = round(mean(loan_amount_000s)))

## Count of property types
owner_occupancy_2 <- data %>%
  group_by(owner_occupancy_name) %>%
  count()

## Merging together
owner_occupancy <- merge(owner_occupancy_1,owner_occupancy_2)
owner_occupancy
```

```{r}
## Summarize loan amount by loan purpose
loan_purpose_1 <- data %>%
  group_by(loan_purpose_name) %>%
  summarize(mean_loan_amount_000s = round(mean(loan_amount_000s)))

## Count of property types
loan_purpose_2 <- data %>%
  group_by(loan_purpose_name) %>%
  count()

## Merging together
loan_purpose <- merge(loan_purpose_1,loan_purpose_2)
loan_purpose
```

```{r}
## Histogram
ggplot(data = loan_purpose, aes(x = loan_purpose_name, y = n)) + 
    geom_col(fill = "skyblue3") +
    geom_text(aes(label = n), size = 4, vjust= -0.5) +
    ylim(0,16000) +
    ggtitle("Number of Loans by Loan Purpose\n") +
    xlab("\nLoan Purpose") +
    ylab("Number of Loans\n")
```

```{r}
## Making a simple bar chart to show disparities
ggplot(data = loan_purpose, aes(x = loan_purpose_name, y = mean_loan_amount_000s)) +
    geom_col(fill = "skyblue3") +
    geom_text(aes(label = mean_loan_amount_000s), size = 3.5, vjust= -0.5) +
    ylim(0,600) +
    ggtitle("Average Loan Amount by Loan Purpose\n") +
    xlab("\nLoan Purpose") +
    ylab("Average Loan Amount (Thousands)\n")
```

## Section 3: Condensing the Various Race/Ethnicity Fields to One Measure per Application

```{r}
## Look at summaries for race/ethnicity variables
as.data.frame(table(data$applicant_ethnicity_name))
as.data.frame(table(data$applicant_race_name_1))
as.data.frame(table(data$applicant_race_name_2))
```


```{r}
## Condensing race fields to one measure
data$applicant_race_combined <- ifelse(data$applicant_race_name_2 == "",data$applicant_race_name_1,"Multiple Races")

as.data.frame(table(data$applicant_race_combined))
```

```{r}
## Condensing race and ethnicity fields to one measure (long)
data$applicant_race_ethnicity_long <- paste(data$applicant_ethnicity_name,data$applicant_race_combined, sep = " / ")

as.data.frame(table(data$applicant_race_ethnicity_long))
```

```{r}
## Condensing race and ethnicity fields to one measure
data$applicant_race_ethnicity <- ifelse(data$applicant_ethnicity_name == "Hispanic or Latino", data$applicant_ethnicity_name,data$applicant_race_combined)

as.data.frame(table(data$applicant_race_ethnicity))
```

```{r}
## Cleaning up some of the categories above
data$applicant_race_ethnicity[data$applicant_race_ethnicity == "Information not provided by applicant in mail, Internet, or telephone application"] <- "Unknown"
data$applicant_race_ethnicity[data$applicant_race_ethnicity == "Not applicable"] <- "Unknown"

as.data.frame(table(data$applicant_race_ethnicity))
```

```{r}
## Look at summaries for race/ethnicity variables for co-applicant
as.data.frame(table(data$co_applicant_ethnicity_name))
as.data.frame(table(data$co_applicant_race_name_1))
as.data.frame(table(data$co_applicant_race_name_2))
```

```{r}
## Condensing race fields to one measure for co-applicant
data$co_applicant_race_combined <- ifelse(data$co_applicant_race_name_2 == "",data$co_applicant_race_name_1,"Multiple Races")

as.data.frame(table(data$co_applicant_race_combined))
```

```{r}
## Condensing race and ethnicity fields to one measure for co-applicant (long)
data$co_applicant_race_ethnicity_long <- paste(data$co_applicant_ethnicity_name,data$co_applicant_race_combined, sep = " / ")

as.data.frame(table(data$co_applicant_race_ethnicity_long))
```

```{r}
## Condensing race and ethnicity fields to one measure
data$co_applicant_race_ethnicity <- ifelse(data$co_applicant_ethnicity_name == "Hispanic or Latino", data$co_applicant_ethnicity_name,data$co_applicant_race_combined)

as.data.frame(table(data$co_applicant_race_ethnicity))
```

```{r}
## Cleaning up some of the categories above
data$co_applicant_race_ethnicity[data$co_applicant_race_ethnicity == "Information not provided by applicant in mail, Internet, or telephone application"] <- "Unknown"
data$co_applicant_race_ethnicity[data$co_applicant_race_ethnicity == "Not applicable"] <- "Unknown"

as.data.frame(table(data$co_applicant_race_ethnicity))
```

```{r}
## Creating a combined applicants column for race and ethnicity (long)
data$applicants_combined_race_ethnicity_long <- paste(data$applicant_race_ethnicity,data$co_applicant_race_ethnicity, sep = " / ")

as.data.frame(table(data$applicants_combined_race_ethnicity_long))
```

```{r}
## Creating a combined applicants column for race and ethnicity
data$applicants_combined_race_ethnicity <- ifelse(data$co_applicant_race_ethnicity == "No co-applicant", data$applicant_race_ethnicity, ifelse(data$applicant_race_ethnicity == data$co_applicant_race_ethnicity, data$applicant_race_ethnicity, ifelse(data$applicant_race_ethnicity == "Unknown", "Unknown", ifelse(data$co_applicant_race_ethnicity == "Unknown", "Unknown", "Multiple Races"))))

as.data.frame(table(data$applicant_race_name_1))
as.data.frame(table(data$applicant_race_ethnicity))
as.data.frame(table(data$applicants_combined_race_ethnicity))
```

## Section 4: Condensing Sex Fields to One Measure per Application

```{r}
## Look at summaries for sex variables
as.data.frame(table(data$applicant_sex_name))
as.data.frame(table(data$co_applicant_sex_name))
```

```{r}
## Cleaning up extraneous values
data$applicant_sex_name[data$applicant_sex_name == "Information not provided by applicant in mail, Internet, or telephone application"] <- "Unknown"
data$applicant_sex_name[data$applicant_sex_name == "Not applicable"] <- "Unknown"

data$co_applicant_sex_name[data$co_applicant_sex_name == "Information not provided by applicant in mail, Internet, or telephone application"] <- "Unknown"
data$co_applicant_sex_name[data$co_applicant_sex_name == "Not applicable"] <- "Unknown"

as.data.frame(table(data$applicant_sex_name))
as.data.frame(table(data$co_applicant_sex_name))
```

```{r}
## Creating combined variable
data$applicants_combined_sex <- paste(data$applicant_sex_name,data$co_applicant_sex_name, sep = " / ")

as.data.frame(table(data$applicants_combined_sex))
```

```{r}
## Cleaning up all unknowns
data$applicants_combined_sex[data$applicants_combined_sex == "Female / Unknown"] <- "Unknown"
data$applicants_combined_sex[data$applicants_combined_sex == "Male / Unknown"] <- "Unknown"
data$applicants_combined_sex[data$applicants_combined_sex == "Unknown / Female"] <- "Unknown"
data$applicants_combined_sex[data$applicants_combined_sex == "Unknown / Male"] <- "Unknown"
data$applicants_combined_sex[data$applicants_combined_sex == "Unknown / No co-applicant"] <- "Unknown"
data$applicants_combined_sex[data$applicants_combined_sex == "Unknown / Unknown"] <- "Unknown"

as.data.frame(table(data$applicants_combined_sex))
```

## Section 5: Developing the 1st Publication Visualization

```{r}
## Check out racial/ethnicity differences by loan purpose
data %>%
  group_by(loan_purpose_name, applicants_combined_race_ethnicity) %>%
  summarize(mean_loan_amount_000s = round(mean(loan_amount_000s)))
```

```{r}
## Filter down to just home purchases
data_purchases <- data %>%
  filter(loan_purpose_name == "Home purchase")

data_purchases
# now just 15,368 observations
```

```{r}
## Dropping the "Unknown" race/ethnicity values, since race/ethnicity will be the focus of this portion of the analysis
data_purchases_race <- data_purchases %>%
  filter(applicants_combined_race_ethnicity != "Unknown")

data_purchases_race
# now just 10,079 observations
```


```{r}
## Check out racial/ethnicity differences for home purchases
## Summarize loan amount by race/ethnicity
race_ethnicity_1 <- data_purchases_race %>%
  group_by(applicants_combined_race_ethnicity) %>%
  summarize(mean_loan_amount_000s = round(mean(loan_amount_000s)))

## Count by race/ethnicity
race_ethnicity_2 <- data_purchases_race %>%
  group_by(applicants_combined_race_ethnicity) %>%
  count()

## Merging together
race_ethnicity <- merge(race_ethnicity_1,race_ethnicity_2)

## Percentages
race_ethnicity <- race_ethnicity %>%
  mutate(percentage = n / nrow(data_purchases_race))

race_ethnicity
```

```{r}
## Let's read in some demographic data on DC to compare the above with
# Source: https://data.census.gov/cedsci/table?g=0400000US11&tid=ACSDT1Y2017.B03002&hidePreview=false
dc_demographic <- read.csv("Data_Inputs/dc_demographic_data.csv", stringsAsFactors = FALSE, skip = 1)
dc_demographic <- as_tibble(dc_demographic)

#dc_demographic
```

```{r}
## Cleaning it up
dc_demographic <- dc_demographic %>%
  select(-NAME) %>%
  filter(Geographic.Area.Name == "Estimate!!Total" |
         Geographic.Area.Name == "Estimate!!Total!!Not Hispanic or Latino!!White alone" |
         Geographic.Area.Name == "Estimate!!Total!!Not Hispanic or Latino!!Black or African American alone" |  
         Geographic.Area.Name == "Estimate!!Total!!Not Hispanic or Latino!!American Indian and Alaska Native alone" |
         Geographic.Area.Name == "Estimate!!Total!!Not Hispanic or Latino!!Asian alone" |
         Geographic.Area.Name == "Estimate!!Total!!Not Hispanic or Latino!!Native Hawaiian and Other Pacific Islander alone" |
         Geographic.Area.Name == "Estimate!!Total!!Not Hispanic or Latino!!Some other race alone" |
         Geographic.Area.Name == "Estimate!!Total!!Not Hispanic or Latino!!Two or more races" |
         Geographic.Area.Name == "Estimate!!Total!!Hispanic or Latino")

dc_demographic
```

```{r}
## Making the labels match with our other data
dc_demographic$Geographic.Area.Name[dc_demographic$Geographic.Area.Name == "Estimate!!Total"] = "Total"

dc_demographic$Geographic.Area.Name[dc_demographic$Geographic.Area.Name == "Estimate!!Total!!Not Hispanic or Latino!!White alone"] = "White"

dc_demographic$Geographic.Area.Name[dc_demographic$Geographic.Area.Name == "Estimate!!Total!!Not Hispanic or Latino!!Black or African American alone"] = "Black or African American"

dc_demographic$Geographic.Area.Name[dc_demographic$Geographic.Area.Name == "Estimate!!Total!!Not Hispanic or Latino!!American Indian and Alaska Native alone"] = "American Indian or Alaska Native"

dc_demographic$Geographic.Area.Name[dc_demographic$Geographic.Area.Name == "Estimate!!Total!!Not Hispanic or Latino!!Asian alone"] = "Asian"

dc_demographic$Geographic.Area.Name[dc_demographic$Geographic.Area.Name == "Estimate!!Total!!Not Hispanic or Latino!!Native Hawaiian and Other Pacific Islander alone"] = "Native Hawaiian or Other Pacific Islander"

dc_demographic$Geographic.Area.Name[dc_demographic$Geographic.Area.Name == "Estimate!!Total!!Not Hispanic or Latino!!Some other race alone"] = "Other Race"

dc_demographic$Geographic.Area.Name[dc_demographic$Geographic.Area.Name == "Estimate!!Total!!Not Hispanic or Latino!!Two or more races"] = "Multiple Races"

dc_demographic$Geographic.Area.Name[dc_demographic$Geographic.Area.Name == "Estimate!!Total!!Hispanic or Latino"] = "Hispanic or Latino"

dc_demographic
```

```{r}
## Renaming columns and calculating percentages
dc_demographic <- dc_demographic %>%
  select(race_ethnicity = Geographic.Area.Name, dc_n = District.of.Columbia) %>%
  mutate(dc_n = as.numeric(dc_n)) %>%
  mutate(dc_percentage = dc_n / 693972)

dc_demographic
```

```{r}
## Dropping total and Other Race so the categories line up between the two dfs
dc_demographic %>%
  filter(race_ethnicity != "Total") %>%
  filter(race_ethnicity != "Other Race")
```

```{r}
## Merge the dc data into the df
race_ethnicity <- merge(race_ethnicity, dc_demographic, by.x = "applicants_combined_race_ethnicity", by.y = "race_ethnicity")

race_ethnicity
```

```{r}
## Need to gather the two percentage columns into one
race_ethnicity <- gather(race_ethnicity, percentage_type, percentage_value, percentage, dc_percentage)

race_ethnicity
```

```{r}
## Manipulate dc_percentage values to be negative for graphing purposes
race_ethnicity <- race_ethnicity %>%
  mutate(percentage_value = if_else(percentage_type == "dc_percentage", -percentage_value, percentage_value))

race_ethnicity
```

```{r}
## Dropping to just white, black, hispanic, and asian
race_ethnicity <- race_ethnicity %>%
  filter(applicants_combined_race_ethnicity != "Native Hawaiian or Other Pacific Islander") %>%
  filter(applicants_combined_race_ethnicity != "American Indian or Alaska Native") %>%
  filter(applicants_combined_race_ethnicity != "Multiple Races")

## Converting percentage value column to clean formatted percentages for text labels
race_ethnicity <- race_ethnicity %>%
  mutate(percentage_value_formatted = abs(percentage_value)) %>%
  mutate(percentage_value_formatted = scales::percent(percentage_value_formatted, accuracy = 1L))

race_ethnicity
```

```{r}
## Making a barchart of race/ethnicity mortgage applications
positions <- c("Asian", "Hispanic or Latino", "White", "Black or African American")  ## setting the order

plot1 <- ggplot(race_ethnicity, aes(x = applicants_combined_race_ethnicity, y = percentage_value, group = percentage_type, fill = percentage_type)) +      ## initiating the plot                                       
  geom_bar(stat = "identity") +   ## making it a bar chart
  coord_flip() +  ## flipping it
  scale_x_discrete(limits = positions) +  ## implementing the set order from above
  #ylim(-0.55,0.65)
  scale_y_continuous(breaks = seq(-0.75, 0.75, 0.25), 
                     labels = abs(seq(-0.75, 0.75, 0.25)),
                     limits = c(-0.55,0.65)) +   ## setting the percentage values to my desired values
  geom_text(aes(label = percentage_value_formatted), size = 4.5, hjust=ifelse(race_ethnicity$percentage_value < 0, 1.2, -0.3)) +  ## adding text labels
  scale_fill_manual(values = c("grey50", "greenyellow"),
                    labels = c("Percentage of DC Population",
                               "Percentage of DC Mortgage Applications")) +   ## setting desired colors and modifying legend labels
  labs(x = "Applicants'\nRace/Ethnicity\n",
       y = "\nPercentage",
       title = "\nBlack and Hispanic consumers are under-represented in mortgage\napplications in Washington, D.C.",
       subtitle = "\nThe graph shows the disparities by race and ethnicity between the proportion of the overall population (on the left),\nand the proportion of mortgage applications for new home purchases (on the right), in Washington, D.C. in 2017.\n",
       caption = "\n\nSources: Census ACS (population demographics), Consumer Financial Protection Bureau (HMDA data)\n") +
  theme_light() +
  theme(legend.position = "bottom",
        legend.title = element_blank(),
        plot.title = element_text(size = 20),
        axis.text=element_text(size=12),
        axis.title=element_text(size=12),
        legend.text=element_text(size=12))


plot1
```

```{r}
## Write to png
#ggsave("Vizzes/race_ethnicity_applications.png", plot = plot1, width = 11, height = 6)
```

## Section 6: Developing the 2nd Publication Visualization

```{r}
## Dropping to just white, black, hispanic, and asian
data_purchases_race_2 <- data_purchases_race %>%
  filter(applicants_combined_race_ethnicity != "Native Hawaiian or Other Pacific Islander") %>%
  filter(applicants_combined_race_ethnicity != "American Indian or Alaska Native") %>%
  filter(applicants_combined_race_ethnicity != "Multiple Races")

data_purchases_race_2
# note that we're now down to 9,280 observations
```

```{r}
## Boxplot plot for loan amount by race/ethnicity
positions_2 <- c("Black or African American", "White", "Hispanic or Latino", "Asian")  ## setting the order
plot2 <- ggplot(data_purchases_race_2, aes(x = applicants_combined_race_ethnicity, y = loan_amount_000s, color = applicants_combined_race_ethnicity)) +    ## initiating the plot
    #geom_dotplot(binaxis = "y", stackdir = "center", binwidth = 5) +
    geom_boxplot() +   ## making it a box plot
    scale_x_discrete(limits = positions_2) +  ## implementing the set order from above
    scale_color_manual(values = c("skyblue3","mediumseagreen","mediumpurple3","indianred2")) +   ## setting desired colors
    #ylim(0,2000) +
    scale_y_continuous(labels = comma, limits = c(0,2000)) +  ## setting and formatting y axis labels
    labs(x = "\nApplicants' Race/Ethnicity",
       y = "Loan Amount\n(Thousands of\nDollars)\n",
       title = "\nBlack and Hispanic consumers tend to apply for smaller loans than\nwhite and Asian consumers, on average.",
       subtitle = "\nThe graph shows the distribution of the size of the loans requested for new home purchases by consumers in each race/ethnicity\ngroup in Washington, D.C. in 2017.\n",
       caption = "\n\nSource: Consumer Financial Protection Bureau (HMDA data)\n") +
  theme_light() +
  theme(legend.position = "none",
        legend.title = element_blank(),
        plot.title = element_text(size = 20),
        axis.text=element_text(size=12),
        axis.title=element_text(size=12),
        axis.title.y = element_text(angle = 0, vjust = 0.5))  ## removing legend and formatting text label sizes etc.


plot2
```

```{r}
## Create list of medians for manual text labels
data_purchases_race_2 %>%
  group_by(applicants_combined_race_ethnicity) %>%
  summarise(median(loan_amount_000s))
```

```{r}
## Write to png
#ggsave("Vizzes/race_ethnicity_loan_amount.png", plot = plot2, width = 11, height = 8.5)
```

## Section 7: Developing the 3rd Production Visualization

```{r}
## Looking at outcomes by race/ethnicity
race_ethnicity_outcomes <- data_purchases_race_2 %>%
  group_by(applicants_combined_race_ethnicity) %>%
  count(action_taken_name) %>%
  mutate(percentage = n / sum(n))

race_ethnicity_outcomes
```

```{r}
## Filtering for just the denials
race_ethnicity_denials <- race_ethnicity_outcomes %>%
  filter(action_taken_name == "Application denied by financial institution")

race_ethnicity_denials
```

```{r}
## plotting the denial rates
positions_3 <- c("Black or African American", "White", "Hispanic or Latino", "Asian")  ## setting the order
plot3 <- ggplot(race_ethnicity_denials, aes(x = applicants_combined_race_ethnicity, y = percentage, fill = applicants_combined_race_ethnicity)) +  ## initiating the plot
  geom_col(width = 0.7) +   ## making it a bar chart
  scale_x_discrete(limits = positions_3) +   ## implementing the set order from above
  scale_fill_manual(values = c("skyblue3","mediumseagreen","mediumpurple3","indianred2")) +  ## setting desired colors
  geom_text(aes(label = paste((round(percentage,3)*100),"%", sep = "")), size = 5, vjust = -0.5) +  ## adding text labels
  ylim(0,0.13) +   ## setting y-axis scale
  labs(x = "\nApplicants' Race/Ethnicity",
       y = "Percent Denied\n",
       title = "\nBlack and Hispanic consumers are denied mortgage loans at higher rates\nthan white and Asian consumers.",
       subtitle = "\nThe graph shows the percentage of mortgage applications for new home purchases by consumers in each race/ethnicity group that were\ndenied in Washington, D.C. in 2017.\n",
       caption = "\n\nSource: Consumer Financial Protection Bureau (HMDA data)\n") +
  theme_light() +
  theme(legend.position = "none",
        legend.title = element_blank(),
        plot.title = element_text(size = 20),
        axis.text=element_text(size=12),
        axis.title=element_text(size=12),
        axis.title.y = element_text(angle = 0, vjust = 0.5))  ## removing legend and formatting text label sizes etc

plot3
```

```{r}
## Write to png
#ggsave("Vizzes/race_ethnicity_denials.png", plot = plot3, width = 11, height = 8.5)
```

## Section 8: Filtering and Exporting Data for 4th Production Visualization in Tableau

```{r}
## Checking out denial reasons
denial_reasons <- data_purchases_race_2 %>%
  group_by(applicants_combined_race_ethnicity) %>%
  count(denial_reason_name_1) %>%
  filter(denial_reason_name_1 != "") %>%
  mutate(percentage = n / sum(n))

denial_reasons
```

```{r}
## Write to csv
#write.csv(denial_reasons,'Data_Outputs/denial_reasons.csv')
```

## Section 9: Statistical Modeling on Denials (Didn't pan out to anything used in final project)

```{r}
## Adding denied column in data_purchases_race_2
data_purchases_race_2$denied <- 0
data_purchases_race_2$denied[data_purchases_race_2$action_taken_name == "Application denied by financial institution"] <- 1

## Adding column for loan amount / income
data_purchases_race_2 <- data_purchases_race_2 %>%
  mutate(loan_amount_income_ratio = loan_amount_000s / applicant_income_000s)

data_purchases_race_2
```

```{r}
## Checking on missing values
data_purchases_race_2_na <- data_purchases_race_2 %>%
    summarise_all(funs(sum(is.na(.))))

t(data_purchases_race_2_na)
```

```{r}
## Creating a version of the data with NA's removed for modeling
data_purchases_race_2_logit <- data_purchases_race_2 %>%
  drop_na(loan_amount_income_ratio)

data_purchases_race_2_logit  # now 9,231 rows
```

```{r}
## Running the logit model - just race/ethnicity
logit = glm(denied ~ relevel(factor(applicants_combined_race_ethnicity), ref = "White"), 
         data = data_purchases_race_2_logit, family = binomial(link = "logit"))

summary(logit)
```

```{r}
## Running the logit model - race/ethnicity + loan_amount_income_ratio
logit_2 = glm(denied ~ loan_amount_income_ratio + relevel(factor(applicants_combined_race_ethnicity), ref = "White"), 
         data = data_purchases_race_2_logit, family = binomial(link = "logit"))

summary(logit_2)
```

```{r}
## Getting the fitted values from the 2nd logit model in the dataframe
data_purchases_race_2_logit$fitted_values_logit <- fitted(logit_2)

data_purchases_race_2_logit
```

```{r}
## Quick plot to check relationship
ggplot(data = data_purchases_race_2_logit %>% filter(loan_amount_income_ratio < 20), aes(x = loan_amount_income_ratio, y = fitted_values_logit, color = applicants_combined_race_ethnicity)) +
    geom_point(alpha=0.7, size = 2)
```

## Section 10: Filtering and Exporting Data for 5th and 6th Production Visualizations

```{r}
## Calculating percentage of applications by race/ethnicity by census tract
tract_race_data <- data_purchases_race_2 %>%
  group_by(census_tract_number) %>%
  count(applicants_combined_race_ethnicity) %>%
  mutate(percentage = n / sum(n)) %>%
  select(-n) %>%
  spread(applicants_combined_race_ethnicity,percentage) %>%
  rename(Black_or_African_American = `Black or African American`) %>%
  rename(Hispanic_or_Latino = `Hispanic or Latino`) %>%
  mutate(Asian = replace_na(Asian,0)) %>%
  mutate(Black_or_African_American = replace_na(Black_or_African_American,0)) %>%
  mutate(Hispanic_or_Latino = replace_na(Hispanic_or_Latino,0)) %>%
  mutate(White = replace_na(White,0)) %>%
  drop_na(census_tract_number)

## Getting the minority pop values by tract
minority_pop <- data_purchases_race_2 %>%
  group_by(census_tract_number) %>%
  count(minority_population)

## Merging together
tract_race_data <- merge(tract_race_data, minority_pop)

tract_race_data
```

```{r}
## Re-formatting so race/ethnicity values are in the same column and standardizing the scales/rounding
tract_race_data_2 <- tract_race_data %>%
  select(-Asian,-Hispanic_or_Latino) %>%
  gather(race_ethnicity,percentage_applications,Black_or_African_American:White) %>%
  arrange(census_tract_number) %>%
  mutate(percentage_applications = round(percentage_applications * 100)) %>%
  mutate(minority_population = round(minority_population))

tract_race_data_2
```

```{r}
## Plotting applications vs. tract demographics - ended up recreating this in Tableau with dynamic functionality instead
plot4 <- ggplot(data = tract_race_data_2 %>% filter(n >20), aes(x = minority_population, y = percentage_applications, color = race_ethnicity)) +
    geom_point(alpha=0.7, size = 2) +
    scale_color_manual(values = c("mediumseagreen","indianred2"),
                       labels = c("Black or African-American",
                               "White")) +
    labs(x = "\nCensus Tract Non-White Population Percentage",
       y = "Percentage of Applications\nby Race/Ethnicity\n",
       title = "\nWhite consumers make up a majority of mortgage applications in most census\ntracts, but applications drop off significantly in tracts with large non-White\npopulations.",
       subtitle = "\nThe graph shows the relationship between the percentage of mortgage applications for new home purchases by White and Black consumers in\neach census tract, and the demographic makeup of the census tract, in Washington, D.C. in 2017. \n",
       caption = "\n\nSource: Consumer Financial Protection Bureau (HMDA data on mortgage apllications)\n") +
    theme_light() +
    theme(legend.position = "right",
        legend.title = element_blank(),
        plot.title = element_text(size = 20),
        axis.text=element_text(size=12),
        axis.title=element_text(size=12))

plot4
```

```{r}
## Writing dataset to csv
#write_csv(tract_race_data_2, 'Data_Outputs/tract_race_data_2.csv')
```

```{r}
## Altering format of census tract values so they speak to my shape file;  creating delta of white applications and white pop
tract_race_data_3 <- tract_race_data %>%
  mutate(census_tract_integer = as.integer(census_tract_number)) %>%
  mutate(census_tract_detail = str_sub(sprintf("%.2f",census_tract_number),-2)) %>%
  mutate(census_tract_leading_0 = formatC(census_tract_integer, width = 4, format = "d", flag = "0")) %>%
  mutate(census_tract_id = paste0(census_tract_leading_0,census_tract_detail)) %>%
  select(-census_tract_integer,-census_tract_detail,-census_tract_leading_0) %>%
  mutate(white_population = 100 - minority_population) %>%
  mutate(white_application_percentage = White * 100) %>%
  mutate(white_application_population_delta = white_application_percentage - white_population)

tract_race_data_3
```

```{r}
## Writing dataset to csv
#write_csv(tract_race_data_3, 'Data_Outputs/tract_race_data_3.csv')
```

## Section 11: Exploring Income Data Further (Didn't pan out to anything used in final project)

```{r}
## Calculating mean/median income value by census tract
tract_income_data <- data_purchases_race_2 %>%
  group_by(census_tract_number) %>%
  summarise(mean_applicant_income = mean(applicant_income_000s, na.rm = TRUE), median_applicant_income = median(applicant_income_000s, na.rm = TRUE)) %>%
  drop_na(census_tract_number)

## Getting the tract to msamd income
tract_msamd <- data_purchases_race_2 %>%
  group_by(census_tract_number) %>%
  count(tract_to_msamd_income)

## Merging together
tract_income_data <- merge(tract_income_data, tract_msamd)

tract_income_data
```

```{r}
## Quick plot to check relationship
ggplot(data = tract_income_data, aes(x = tract_to_msamd_income, y = median_applicant_income)) +
    geom_point(alpha=0.7, size = 2)
```

```{r}
## Calculating mean/median income value by census tract
tract_loan_value <- data_purchases_race_2 %>%
  group_by(census_tract_number) %>%
  summarise(mean_loan_value = mean(loan_amount_000s, na.rm = TRUE), median_loan_value = median(loan_amount_000s, na.rm = TRUE)) %>%
  drop_na(census_tract_number)

## Getting the tract to msamd income
tract_msamd <- data_purchases_race_2 %>%
  group_by(census_tract_number) %>%
  count(tract_to_msamd_income)

## Merging together
tract_loan_value <- merge(tract_loan_value, tract_msamd)

tract_loan_value
```

```{r}
## Quick plot to check relationship
ggplot(data = tract_loan_value, aes(x = tract_to_msamd_income, y = mean_loan_value)) +
    geom_point(alpha=0.7, size = 2)
```

